---
title: "Linking genetic diversity to range shift dynamics"
output: word_document
date: "2026-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
#required packages
list.of.packages <- c(
  #"doParallel", "parallel","foreach",
  #"pbapply",
  "dplyr", "tidyr", #"data.table",
  #"scales","effects","psych", 
  "glmmTMB", "lme4", "lmerTest",#"here","rlist","partR2",
  #"ggtext","gridExtra","grid","lattice","viridis",
  "performance","MuMIn")#,"glmm.hp") 
#
#
#new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
#
#if(length(new.packages)) install.packages(new.packages)
#
sapply(list.of.packages, require, character.only = TRUE)

###########################
#define the data repository
#if(!dir.exists(here("Output/full_model"))){
#  dir.create(here("Output/full_model"),recursive = TRUE)
#}
#
#dir.in=here("Data") #to change accordingly to the location of the data
#dir.out=here("Output/full_model") #to change. It's the repository where the results are saved

# Load data
#mydataset <- read.csv2(here(dir.in,"gen_data_final_fonseca.csv"),
#                       sep=",",dec=".",h=T) 


```

```{r dataselection, message = FALSE, warning=FALSE}
# load data
mydataset <- read.csv('data/gen_data_final_fonseca.csv')

#Data selection
## Latitude data
mydatatogo <- mydataset  %>%
  dplyr::filter(Type == "LAT",
                shift_vel_sign == "pospos" | shift_vel_sign == "negneg", # Select only shifts in the same direction of velocity
                SHIFT != 0, # remove non-significant shifts
                # Nucleotide_diversity > 0 # select only GD values > 0
  ) %>% 
  mutate(
    # Climate velocity
    vel = as.numeric(velocity),
    vel_abs = abs(vel),
    vel_abs_log = log(vel_abs),
    vel_abs_log1p = log1p(vel_abs),
    
    # Shift
    SHIFT = SHIFT, # to deal with zero shift
    SHIFT_abs = abs(SHIFT),
    SHIFT_abs_log = log(SHIFT_abs),
    SHIFT_abs_log1p = log1p(SHIFT_abs),
    
    # Genetic diversity
    GD = Nucleotide_diversity, 
    GD_sqrt = sqrt(GD),
    GD_log = log(GD),
    GD_log1p = log1p(GD),
    
    # Methods
    Lat = abs(Latitude),
    Lat_band = round(Lat,0),
    ID.area = scale(ID.area),
    DUR = scale(DUR),
    LogExtent = log(Extent),
    START = scale(START),
    Param = factor(Param),
    Group = factor(Group),
    spp = factor(spp), 
    ExtentF = cut(Extent,
                  ordered_result = TRUE,
                  breaks=seq(min(Extent), max(Extent), length.out=10),
                  include.lowest=TRUE),
    NtempUnitsF = cut(NtempUnits,
                      ordered_result = TRUE,
                      breaks=seq(min(NtempUnits), max(NtempUnits), length.out=10),
                      include.lowest=TRUE)) %>%
  
  dplyr::select(
    # Genetic diversity
    GD, GD_log, GD_log1p, GD_sqrt, TajimasD,
    # Shift
    SHIFT, SHIFT_abs, SHIFT_abs_log, SHIFT_abs_log1p, 
    # SHIFT_cor, SHIFT_cor_abs, SHIFT_cor_abs_log, SHIFT_cor_raw, SHIFT_abs_log_scale,
    # Velocity
    vel, vel_abs, vel_abs_log, vel_abs_log1p, 
    trend.mean,
    # Methods + Taxonomy
    Article_ID, 
    Hemisphere,
    shift_vel_sign,
    Lat, # latitudinal position where GD was collected
    Lat_band,
    DUR, Nperiodes, LogNtempUnits, NtempUnits, Extent, LogExtent, ContinuousGrain, Quality, PrAb, ExtentF, NtempUnitsF,
    Param, Group, spp, Class, Order, Family, Genus, 
    ECO, Uncertainty_Parameter, Uncertainty_Distribution, Grain_size, Data, Article_ID
  ) 

# transform continuous variables
cont_vars <- c(1:14,18, 20:26)

mydatatogo[,cont_vars] <- lapply(mydatatogo[,cont_vars], as.numeric)
mydatatogo[,-cont_vars] <- lapply(mydatatogo[,-cont_vars], function(x) factor(x, levels = unique(x)))

##eliminating species with uncertain obs
mydatatogo=subset(mydatatogo,spp!="Agrilus planipennis")
mydatatogo=subset(mydatatogo,spp!="Chrysodeixis eriosoma")
mydatatogo=droplevels(mydatatogo)

## Filter Classes with at least 5 species per Param
n_sps = 10

test <- mydatatogo %>%
  group_by(Class,Param) %>%
  summarise(N_spp = length(unique(spp))) %>% # how many species per parameter?
  dplyr::filter(N_spp >= n_sps) # select classes with > n_sps per param

test <- mutate(test, Class_Param = paste(Class, Param))

mydatatogo <- mydatatogo %>%
  mutate(Class_Param = paste(Class, Param)) %>%
  dplyr::filter(Class_Param %in% test$Class_Param) %>%
  dplyr::select(-Class_Param)

mydatatogo[,-cont_vars] <- lapply(mydatatogo[,-cont_vars], function(x) factor(x, levels = unique(x)))

## Extra fixes
# Set the reference param level to the centroid of species obs
mydatatogo$Param <- relevel(mydatatogo$Param, ref = "O") 

# Taxonomy data
taxatable <- mydatatogo %>%
  group_by(Class, Order, Family) %>%
  summarise(Species = length(unique(spp)),
            Shift = n())

#write.csv(taxatable,
#          here("Output/taxa_table.csv"),
#          row.names = FALSE)

```


## Model weights

```{r weights}
#### Variance partitioning #####

mydatatogo2=mydatatogo
mydatatogoCE=subset(mydatatogo2,Param=="O")
x1=data.frame(table(mydatatogoCE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoCE=merge(mydatatogoCE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoLE=subset(mydatatogo2,Param=="LE")
x1=data.frame(table(mydatatogoLE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoLE=merge(mydatatogoLE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoTE=subset(mydatatogo2,Param=="TE")
x1=data.frame(table(mydatatogoTE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoTE=merge(mydatatogoTE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoTE$weight_obs_spp=mydatatogoTE$weight_obs_spp*(1/3)
mydatatogoCE$weight_obs_spp=mydatatogoCE$weight_obs_spp*(1/3)
mydatatogoLE$weight_obs_spp=mydatatogoLE$weight_obs_spp*(1/3)

mydatatogo2=rbind(mydatatogoTE,mydatatogoCE,mydatatogoLE)

mydatatogo2$vel_abs_s <- scale(mydatatogo2$vel_abs)
mydatatogo2$GD_s <- scale(mydatatogo2$GD)
```

## Model weights sum to 1:
```{r weightsum}
sum(mydatatogo2$weight_obs_spp)
```

## Fit model:

```{r model}
gam_velxGDxedge1 <- as.formula(
  "SHIFT_abs ~ vel_abs_s + GD_s + Param +
   vel_abs_s:GD_s + vel_abs_s:Param + GD_s:Param +
   vel_abs_s:GD_s:Param +
   LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality +
   (1|Class)")

full <- glmmTMB(gam_velxGDxedge1,
                family = Gamma(link = 'log'),
                weights = mydatatogo2$weight_obs_spp,
                REML = FALSE,
                data = mydatatogo2)

        
```
### No significant effects:
```{r summaryfull}
summary(full) 
```

### Model fit is singular, and random effect variance is approximately 0:
```{r checksing}
check_singularity(full)
```


### Because of this, R2m = R2c (no proportion of variance is attributed to the random effect of class)
```{r r2}
MuMIn::r.squaredGLMM(full)
```


## Bootstrapping
We fit 1000 iterations as an example (resulting R object can be loaded directly in next chunk).

```{r bootstrap, eval=FALSE}
####Fit a single model with Param*GD*climate velocity####

#Obs are weighted to correct for imbalance in obs position (TE,CE,LE) and species frequency
#A GLMM model are fitted to control for method heterogeneity that assessed species range shifts and phylogeny proximity
#The model are bootstrapped to compute ccurate mean, CI95% and significance of effect size.
gam_velxGDxedge1 <- as.formula(
    "SHIFT_abs ~ scale(vel_abs) * scale(GD) * Param + 
    LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality + 
    (1|Class)")
s1=69 #the initial random seed
nB=1000 #the number of bootstraps
#nCPU=10

#setting a local cluster
#my.cluster2 <- parallel::makeCluster(
#    nCPU, 
#    type = "PSOCK"
#)
#clusterExport(cl = my.cluster2, varlist = c("s1","nB","glmmTMB","mydatatogo","gam_velxGDxedge1","ranef","check_singularity"))
#
##register cluster
#doParallel::registerDoParallel(cl = my.cluster2)

ex=lapply(1:nB, function(i) {
    
    gc(reset=T)
    #print(i)
    set.seed(s1+i)
    
    n1=sample(1:nrow(mydatatogo),rep=T,size=nrow(mydatatogo))
    mydatatogo2=mydatatogo[n1,]
    mydatatogoCE=subset(mydatatogo2,Param=="O")
    x1=data.frame(table(mydatatogoCE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoCE=merge(mydatatogoCE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoLE=subset(mydatatogo2,Param=="LE")
    x1=data.frame(table(mydatatogoLE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoLE=merge(mydatatogoLE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoTE=subset(mydatatogo2,Param=="TE")
    x1=data.frame(table(mydatatogoTE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoTE=merge(mydatatogoTE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoTE$weight_obs_spp=mydatatogoTE$weight_obs_spp*(1/3)
    mydatatogoCE$weight_obs_spp=mydatatogoCE$weight_obs_spp*(1/3)
    mydatatogoLE$weight_obs_spp=mydatatogoLE$weight_obs_spp*(1/3)
    
    mydatatogo2=rbind(mydatatogoTE,mydatatogoCE,mydatatogoLE)
    
    gamX1 <- glmmTMB(gam_velxGDxedge1, 
                     family = Gamma(link = "log"),
                     weights = weight_obs_spp,
                     REML=F,
                     data = mydatatogo2)
    round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=data.frame(r2m=r2[2,1],
                  r2c=r2[2,2],
                  moy_GD=mean(mydatatogo2$GD), 
                  sd_GD=sd(mydatatogo2$GD),
                  moy_VC=mean(mydatatogo2$vel_abs), 
                  sd_VC=sd(mydatatogo2$vel_abs),
                  singu=check_singularity(gamX1)[[1]],
                  nB=i)
    
    coeff=data.frame(summary(gamX1)$coeff$cond,
                     var=row.names(summary(gamX1)$coeff$cond),
                     nB=i)
    
    randEff=as.data.frame(ranef(gamX1,condVar=T))
    randEff$nB=i
    
    return(list(r2,coeff,randEff))
    
})

#parallel::stopCluster(cl = my.cluster2)

```


## Summary of bootstrap results
```{r summbootstrap}
ex = readRDS('1k_boot.rds')

nom="allW"
randEff=rlist::list.rbind(lapply(ex,"[[",3))
coeff=rlist::list.rbind(lapply(ex,"[[",2))
r2=rlist::list.rbind(lapply(ex,"[[",1))

   
dir.create('model_out')
dir.out = 'model_out/'
write.table(r2,
            paste0(dir.out, "R2_",nom,".csv"),
            sep=";",dec=".",row=F)
write.table(coeff,
            paste0(dir.out, "coeff_",nom,".csv"),
            sep=";",dec=".",row=F)
write.table(randEff,
            paste0(dir.out, "randEff_",nom,".csv"),
            sep=";",dec=".",row=F)
```

## Singular fit
Model fit is singular, and R2m = R2c for all bootstrap iterations:
```{r sing}
head(r2)

## All iterations have singular fit --------
summary(r2$singu) 

## R2m = R2c for all iterations --------
all.equal(r2$r2m, r2$r2c) 

```


## Analysis of bootstrap output
Summary of model output (res_ok), write results files:  
- summary of R2 for iterations: R2_allW_sel1000.csv  
- summary of R2 overall: summary_R2.csv  
- summary of model coefficients after bootstrapping: summary_coeff.csv  

```{r bootstrapanalysis}
#analysis of raw bootstrap output
test1=function(x,mu=0){
    return(1-(length(x[x>mu])/length(x)))
}

#bootstrap test: H0= coefficient equal to 0; H1= coefficient lower than 0
test2=function(x,mu=0){
    return(1-(length(x[x<mu])/length(x)))
}


mod='allW'
a=1
s1=10 #the seed to make reproducible random stuff
nB=1000
setwd(dir.out)

for(i in 1:length(mod)){
    print(a)
    r2=read.csv2(paste0("R2_",mod[i],".csv"),sep=";",dec=".")
    #r2$nB=1:nrow(r2)
    r2a=subset(r2,is.na(r2m)==F)
    print(paste(mod[i],': ',nrow(r2a),sep=""))
    if(nrow(r2a)>=nB){ #random selection of nB bootstrap model (criteria: model convergence)
        set.seed(s1)
        r2a=r2a[sample(1:nrow(r2a),size=nB,replace=F),]
    }
    if(nrow(r2a)<nB){
        print(paste0("sampling size is less than ",nB," bootstraps"))
    }else{
        resR2=data.frame(type="r2m",moy=mean(r2a$r2m),sd=sd(r2a$r2m),median=median(r2a$r2m),q025=quantile(r2a$r2m,probs=0.025),q975=quantile(r2a$r2m,probs=0.975))
        resR2=rbind(resR2,data.frame(type="r2c",moy=mean(r2a$r2c),sd=sd(r2a$r2c),median=median(r2a$r2c),q025=quantile(r2a$r2c,probs=0.025),q975=quantile(r2a$r2c,probs=0.975)))
        #resR2$n_sing=nrow(subset(r2a,sing==T))
        resR2$model=mod[i]
        
        c1=read.csv2(paste("coeff_",mod[i],".csv",sep=""),sep=";",dec=".",h=T)
        #c1$nB=rep(1:nrow(r2),each=nrow(c1)/6000)
        c1=merge(c1,data.frame(nB=r2a$nB),by.x="nB",by.y="nB")
        rr3=c1
        
        res=data.frame(var=names(tapply(rr3$Estimate,rr3$var,mean)),moy=tapply(rr3$Estimate,rr3$var,mean),sd=tapply(rr3$Estimate,rr3$var,sd),median=tapply(rr3$Estimate,rr3$var,median),q025=tapply(rr3$Estimate,rr3$var,quantile,probs=0.025),
                       p975=tapply(rr3$Estimate,rr3$var,quantile,probs=0.975), pv.inf0=tapply(rr3$Estimate,rr3$var,test2,mu=0),pv.sup0=tapply(rr3$Estimate,rr3$var,test1,mu=0))
        res$model=mod[i]
        if(a==1){
            resR2_ok=resR2
            res_ok=res
        }
        else{
            resR2_ok=rbind(resR2_ok,resR2)
            res_ok=rbind(res_ok,res)
        }
        
        a=a+1
    }
}

write.table(r2a,
            paste0("R2_",nom,"_sel1000.csv"),
            sep=";",dec=".",row=F) #999 selected bootstrapped models

write.table(resR2_ok,
            paste0("summary_R2.csv"),
            sep=";",dec=".",row=F) 

#R2 statistic computed from 5000 boostrap model:
#type= R2 type (r2m= marginal R2; r2c= conditional R2)
#moy= mean R2 value
#sd= R2 standard deviation value
#median= median R2 value
#q025= 2.5% quantile of the bootstrap distribution
#q975= 97.5% quantile of the bootstrap distribution
#n_sing= number of singular model
#model= model name abbreviation

write.table(res_ok,
            ("summary_coeff.csv"),
            sep=";",dec=".",row=F) 
#coefficient statistics
#Var =  Term of the model
#moy = mean effect computed from the 5000 bootstrap model
#sd = standard deviation computed from the 5000 bootstrap model
#med = median effect computed from the 5000 bootstrap model
#q025 = 2.5% quantile of the bootstrap coefficient distribution
#q975 = 97.5% quantile of the bootstrap coefficient distribution
#pv.sup0 = p-value of the bootstrap test testing the significance of the effect (H0= coefficient equal to 0; H1= coefficient greater than 0)
#pv.inf0 = p-value of the bootstrap test testing the significance of the effect (H0= coefficient equal to 0; H1= coefficient lower than 0)
#model= model name abbreviation

```

## Bootstrap summary
There is no statistical support for the 3-way interaction between genetic diversity, climate velocity, and range position.
```{r modres}

# Summary of R2:
resR2_ok

# Model coefficients:
rownames(res_ok)=NULL
res_ok
```


## Adjusted model

We fit the same model nesting a species random effect within class, removing weights and without bootstrapping:

```{r altmodel}

gam_velxGDxedge1.alt <- as.formula(
    "SHIFT_abs ~ scale(vel_abs) * scale(GD) * Param + 
    LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality + 
    (1|Class/spp)")

alt.gam <- glmmTMB(gam_velxGDxedge1.alt, 
                     family = Gamma(link = "log"),
                     #weights = weight_obs_spp,
                     REML=F,
                     data = mydatatogo2)

```


There is no support for an effect of genetic diversity when accounting for class and species:
```{r altsummary}

summary(alt.gam)
```



This lack of effect holds when fitting the original model with a class random effect  without weights or bootstrapping, particularly for the three-way interaction effect:
```{r model_class}
gamX1.cl <- glmmTMB(SHIFT_abs ~ scale(vel_abs) * scale(GD) * Param + 
                      LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality + 
                      (1|Class), 
                   family = Gamma(link = "log"),
                   #weights = weight_obs_spp,
                   REML=F,
                   data = mydatatogo2)

summary(gamX1.cl)
```

