---
title: "Linking genetic diversity to range shift dynamics"
output: word_document
date: "2026-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
#required packages
list.of.packages <- c(
  #"doParallel","foreach",
  #"pbapply",
  "parallel", "dplyr", "tidyr", #"data.table",
  #"scales","effects","psych", 
  "glmmTMB", "lme4", "lmerTest","here",#"rlist","partR2",
  #"ggtext","gridExtra","grid","lattice","viridis",
  "performance","MuMIn")#,"glmm.hp") 
#
#
#new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
#
#if(length(new.packages)) install.packages(new.packages)
#
sapply(list.of.packages, require, character.only = TRUE)

###########################
#define the data repository
#if(!dir.exists(here("Output/full_model"))){
#  dir.create(here("Output/full_model"),recursive = TRUE)
#}
#
#dir.in=here("Data") #to change accordingly to the location of the data
#dir.out=here("Output/full_model") #to change. It's the repository where the results are saved

# Load data
#mydataset <- read.csv2(here(dir.in,"gen_data_final_fonseca.csv"),
#                       sep=",",dec=".",h=T) 


```

```{r dataselection, message = FALSE, warning=FALSE}
# load data
mydataset <- read.csv(here("data", "gen_data_final_fonseca.csv"))

#Data selection
## Latitude data
mydatatogo <- mydataset  %>%
  dplyr::filter(Type == "LAT",
                shift_vel_sign == "pospos" | shift_vel_sign == "negneg", # Select only shifts in the same direction of velocity
                SHIFT != 0, # remove non-significant shifts
                # Nucleotide_diversity > 0 # select only GD values > 0
  ) %>% 
  mutate(
    # Climate velocity
    vel = as.numeric(velocity),
    vel_abs = abs(vel),
    vel_abs_log = log(vel_abs),
    vel_abs_log1p = log1p(vel_abs),
    
    # Shift
    SHIFT = SHIFT, # to deal with zero shift
    SHIFT_abs = abs(SHIFT),
    SHIFT_abs_log = log(SHIFT_abs),
    SHIFT_abs_log1p = log1p(SHIFT_abs),
    
    # Genetic diversity
    GD = Nucleotide_diversity, 
    GD_sqrt = sqrt(GD),
    GD_log = log(GD),
    GD_log1p = log1p(GD),
    
    # Methods
    Lat = abs(Latitude),
    Lat_band = round(Lat,0),
    ID.area = scale(ID.area),
    DUR = scale(DUR),
    LogExtent = log(Extent),
    START = scale(START),
    Param = factor(Param),
    Group = factor(Group),
    spp = factor(spp), 
    ExtentF = cut(Extent,
                  ordered_result = TRUE,
                  breaks=seq(min(Extent), max(Extent), length.out=10),
                  include.lowest=TRUE),
    NtempUnitsF = cut(NtempUnits,
                      ordered_result = TRUE,
                      breaks=seq(min(NtempUnits), max(NtempUnits), length.out=10),
                      include.lowest=TRUE)) %>%
  
  dplyr::select(
    # Genetic diversity
    GD, GD_log, GD_log1p, GD_sqrt, TajimasD,
    # Shift
    SHIFT, SHIFT_abs, SHIFT_abs_log, SHIFT_abs_log1p, 
    # SHIFT_cor, SHIFT_cor_abs, SHIFT_cor_abs_log, SHIFT_cor_raw, SHIFT_abs_log_scale,
    # Velocity
    vel, vel_abs, vel_abs_log, vel_abs_log1p, 
    trend.mean,
    # Methods + Taxonomy
    Article_ID, 
    Hemisphere,
    shift_vel_sign,
    Lat, # latitudinal position where GD was collected
    Lat_band,
    DUR, Nperiodes, LogNtempUnits, NtempUnits, Extent, LogExtent, ContinuousGrain, Quality, PrAb, ExtentF, NtempUnitsF,
    Param, Group, spp, Class, Order, Family, Genus, 
    ECO, Uncertainty_Parameter, Uncertainty_Distribution, Grain_size, Data, Article_ID
  ) 

# transform continuous variables
cont_vars <- c(1:14,18, 20:26)

mydatatogo[,cont_vars] <- lapply(mydatatogo[,cont_vars], as.numeric)
mydatatogo[,-cont_vars] <- lapply(mydatatogo[,-cont_vars], function(x) factor(x, levels = unique(x)))

##eliminating species with uncertain obs
mydatatogo=subset(mydatatogo,spp!="Agrilus planipennis")
mydatatogo=subset(mydatatogo,spp!="Chrysodeixis eriosoma")
mydatatogo=droplevels(mydatatogo)

## Filter Classes with at least 5 species per Param
n_sps = 10

test <- mydatatogo %>%
  group_by(Class,Param) %>%
  summarise(N_spp = length(unique(spp))) %>% # how many species per parameter?
  dplyr::filter(N_spp >= n_sps) # select classes with > n_sps per param

test <- mutate(test, Class_Param = paste(Class, Param))

mydatatogo <- mydatatogo %>%
  mutate(Class_Param = paste(Class, Param)) %>%
  dplyr::filter(Class_Param %in% test$Class_Param) %>%
  dplyr::select(-Class_Param)

mydatatogo[,-cont_vars] <- lapply(mydatatogo[,-cont_vars], function(x) factor(x, levels = unique(x)))

## Extra fixes
# Set the reference param level to the centroid of species obs
mydatatogo$Param <- relevel(mydatatogo$Param, ref = "O") 

# Taxonomy data
taxatable <- mydatatogo %>%
  group_by(Class, Order, Family) %>%
  summarise(Species = length(unique(spp)),
            Shift = n())

#write.csv(taxatable,
#          here("Output/taxa_table.csv"),
#          row.names = FALSE)

```

## Model weights
```{r weights}
#### Variance partitioning #####

mydatatogo2=mydatatogo
mydatatogoCE=subset(mydatatogo2,Param=="O")
x1=data.frame(table(mydatatogoCE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoCE=merge(mydatatogoCE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoLE=subset(mydatatogo2,Param=="LE")
x1=data.frame(table(mydatatogoLE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoLE=merge(mydatatogoLE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoTE=subset(mydatatogo2,Param=="TE")
x1=data.frame(table(mydatatogoTE$spp))
x1=subset(x1,Freq>0)
x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
mydatatogoTE=merge(mydatatogoTE,x1[,c(1,3)],by.x="spp",by.y="Var1")

mydatatogoTE$weight_obs_spp=mydatatogoTE$weight_obs_spp*(1/3)
mydatatogoCE$weight_obs_spp=mydatatogoCE$weight_obs_spp*(1/3)
mydatatogoLE$weight_obs_spp=mydatatogoLE$weight_obs_spp*(1/3)

mydatatogo2=rbind(mydatatogoTE,mydatatogoCE,mydatatogoLE)

mydatatogo2$vel_abs_s <- scale(mydatatogo2$vel_abs)
mydatatogo2$GD_s <- scale(mydatatogo2$GD)
```

## Model weights sum to 1:
```{r weightsum}
sum(mydatatogo2$weight_obs_spp)
```

## Fit model:
```{r model}
gam_velxGDxedge1 <- as.formula(
  "SHIFT_abs ~ vel_abs_s + GD_s + Param +
   vel_abs_s:GD_s + vel_abs_s:Param + GD_s:Param +
   vel_abs_s:GD_s:Param +
   LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality +
   (1|Class)")

full <- glmmTMB(gam_velxGDxedge1,
                family = Gamma(link = 'log'),
                weights = mydatatogo2$weight_obs_spp,
                REML = FALSE,
                data = mydatatogo2)

        
```

No significant effect, inflated SEs
```{r summaryfull}
summary(full) 
```

Model fit is singular, and random effect variance is approximately 0:
```{r checksing}
check_singularity(full)
```

Because of this, R2m = R2c (no proportion of variance is attributed to the random effect of class)
```{r r2}
MuMIn::r.squaredGLMM(full)
```

## Bootstrapping
We fit 1000 iterations as an example (resulting R object can be loaded directly in next chunk).
```{r bootstrap, eval=FALSE}
####Fit a single model with Param*GD*climate velocity####

#Obs are weighted to correct for imbalance in obs position (TE,CE,LE) and species frequency
#A GLMM model are fitted to control for method heterogeneity that assessed species range shifts and phylogeny proximity
#The model are bootstrapped to compute ccurate mean, CI95% and significance of effect size.
gam_velxGDxedge1 <- as.formula(
    "SHIFT_abs ~ scale(vel_abs) * scale(GD) * Param + 
    LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality + 
    (1|Class)")
s1=69 #the initial random seed
nB=1000 #the number of bootstraps
#nCPU=10

#setting a local cluster
#my.cluster2 <- parallel::makeCluster(
#    nCPU, 
#    type = "PSOCK"
#)
#clusterExport(cl = my.cluster2, varlist = c("s1","nB","glmmTMB","mydatatogo","gam_velxGDxedge1","ranef","check_singularity"))
#
##register cluster
#doParallel::registerDoParallel(cl = my.cluster2)

ex=lapply(1:nB, function(i) {
    
    gc(reset=T)
    #print(i)
    set.seed(s1+i)
    
    n1=sample(1:nrow(mydatatogo),rep=T,size=nrow(mydatatogo))
    mydatatogo2=mydatatogo[n1,]
    mydatatogoCE=subset(mydatatogo2,Param=="O")
    x1=data.frame(table(mydatatogoCE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoCE=merge(mydatatogoCE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoLE=subset(mydatatogo2,Param=="LE")
    x1=data.frame(table(mydatatogoLE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoLE=merge(mydatatogoLE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoTE=subset(mydatatogo2,Param=="TE")
    x1=data.frame(table(mydatatogoTE$spp))
    x1=subset(x1,Freq>0)
    x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    mydatatogoTE=merge(mydatatogoTE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    
    mydatatogoTE$weight_obs_spp=mydatatogoTE$weight_obs_spp*(1/3)
    mydatatogoCE$weight_obs_spp=mydatatogoCE$weight_obs_spp*(1/3)
    mydatatogoLE$weight_obs_spp=mydatatogoLE$weight_obs_spp*(1/3)
    
    mydatatogo2=rbind(mydatatogoTE,mydatatogoCE,mydatatogoLE)
    
    gamX1 <- glmmTMB(gam_velxGDxedge1, 
                     family = Gamma(link = "log"),
                     weights = weight_obs_spp,
                     REML=F,
                     data = mydatatogo2)
    round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=data.frame(r2m=r2[2,1],
                  r2c=r2[2,2],
                  moy_GD=mean(mydatatogo2$GD), 
                  sd_GD=sd(mydatatogo2$GD),
                  moy_VC=mean(mydatatogo2$vel_abs), 
                  sd_VC=sd(mydatatogo2$vel_abs),
                  singu=check_singularity(gamX1)[[1]],
                  nB=i)
    
    coeff=data.frame(summary(gamX1)$coeff$cond,
                     var=row.names(summary(gamX1)$coeff$cond),
                     nB=i)
    
    randEff=as.data.frame(ranef(gamX1,condVar=T))
    randEff$nB=i
    
    return(list(r2,coeff,randEff))
    
})

#parallel::stopCluster(cl = my.cluster2)

```


## Summary of bootstrap results
```{r summbootstrap}
ex = readRDS(here('data', '1k_boot.rds'))

nom="allW"
randEff=rlist::list.rbind(lapply(ex,"[[",3))
coeff=rlist::list.rbind(lapply(ex,"[[",2))
r2=rlist::list.rbind(lapply(ex,"[[",1))

# dir.create('model_out')
# dir.out = 'model_out/'
# write.table(r2,
#             paste0(dir.out, "R2_",nom,".csv"),
#             sep=";",dec=".",row=F)
# write.table(coeff,
#             paste0(dir.out, "coeff_",nom,".csv"),
#             sep=";",dec=".",row=F)
# write.table(randEff,
#             paste0(dir.out, "randEff_",nom,".csv"),
#             sep=";",dec=".",row=F)
```

## Singular fit
Model fit is singular, and R2m = R2c for all bootstrap iterations:
```{r sing}
head(r2)

## All iterations have singular fit --------
summary(r2$singu) 

## R2m = R2c for all iterations --------
all.equal(r2$r2m, r2$r2c) 

```


## Analysis of bootstrap output
```{r bootstrapanalysis}
#analysis of raw bootstrap output
test1=function(x,mu=0){
    return(1-(length(x[x>mu])/length(x)))
}

#bootstrap test: H0= coefficient equal to 0; H1= coefficient lower than 0
test2=function(x,mu=0){
    return(1-(length(x[x<mu])/length(x)))
}


mod='allW'
a=1
s1=10 #the seed to make reproducible random stuff
nB=1000
#setwd(dir.out)

for(i in 1:length(mod)){
    print(a)
    r2=rlist::list.rbind(lapply(ex,"[[",1))
    #r2$nB=1:nrow(r2)
    r2a=subset(r2,is.na(r2m)==F)
    print(paste(mod[i],': ',nrow(r2a),sep=""))
    if(nrow(r2a)>=nB){ #random selection of nB bootstrap model (criteria: model convergence)
        set.seed(s1)
        r2a=r2a[sample(1:nrow(r2a),size=nB,replace=F),]
    }
    if(nrow(r2a)<nB){
        print(paste0("sampling size is less than ",nB," bootstraps"))
    }else{
        resR2=data.frame(type="r2m",moy=mean(r2a$r2m),sd=sd(r2a$r2m),median=median(r2a$r2m),q025=quantile(r2a$r2m,probs=0.025),q975=quantile(r2a$r2m,probs=0.975))
        resR2=rbind(resR2,data.frame(type="r2c",moy=mean(r2a$r2c),sd=sd(r2a$r2c),median=median(r2a$r2c),q025=quantile(r2a$r2c,probs=0.025),q975=quantile(r2a$r2c,probs=0.975)))
        #resR2$n_sing=nrow(subset(r2a,sing==T))
        resR2$model=mod[i]
        
        c1 <- do.call(rbind, lapply(ex, `[[`, 2))
        #c1=read.csv2(paste("coeff_",mod[i],".csv",sep=""),sep=";",dec=".",h=T)
        #c1$nB=rep(1:nrow(r2),each=nrow(c1)/6000)
        c1=merge(c1,data.frame(nB=r2a$nB),by.x="nB",by.y="nB")
        rr3=c1
        
        res=data.frame(var=names(tapply(rr3$Estimate,rr3$var,mean)),moy=tapply(rr3$Estimate,rr3$var,mean),sd=tapply(rr3$Estimate,rr3$var,sd),median=tapply(rr3$Estimate,rr3$var,median),q025=tapply(rr3$Estimate,rr3$var,quantile,probs=0.025),
                       p975=tapply(rr3$Estimate,rr3$var,quantile,probs=0.975), pv.inf0=tapply(rr3$Estimate,rr3$var,test2,mu=0),pv.sup0=tapply(rr3$Estimate,rr3$var,test1,mu=0))
        res$model=mod[i]
        if(a==1){
            resR2_ok=resR2
            res_ok=res
        }
        else{
            resR2_ok=rbind(resR2_ok,resR2)
            res_ok=rbind(res_ok,res)
        }
        
        a=a+1
    }
}

# write.table(r2a,
#             paste0("R2_",nom,"_sel1000.csv"),
#             sep=";",dec=".",row=F) #999 selected bootstrapped models
# 
# write.table(resR2_ok,
#             paste0("summary_R2.csv"),
#             sep=";",dec=".",row=F) 

#R2 statistic computed from 5000 boostrap model:
#type= R2 type (r2m= marginal R2; r2c= conditional R2)
#moy= mean R2 value
#sd= R2 standard deviation value
#median= median R2 value
#q025= 2.5% quantile of the bootstrap distribution
#q975= 97.5% quantile of the bootstrap distribution
#n_sing= number of singular model
#model= model name abbreviation

# write.table(res_ok,
#             ("summary_coeff.csv"),
#             sep=";",dec=".",row=F) 
#coefficient statistics
#Var =  Term of the model
#moy = mean effect computed from the 5000 bootstrap model
#sd = standard deviation computed from the 5000 bootstrap model
#med = median effect computed from the 5000 bootstrap model
#q025 = 2.5% quantile of the bootstrap coefficient distribution
#q975 = 97.5% quantile of the bootstrap coefficient distribution
#pv.sup0 = p-value of the bootstrap test testing the significance of the effect (H0= coefficient equal to 0; H1= coefficient greater than 0)
#pv.inf0 = p-value of the bootstrap test testing the significance of the effect (H0= coefficient equal to 0; H1= coefficient lower than 0)
#model= model name abbreviation

```

## Bootstrap summary
There is no statistical support for the 3-way interaction between genetic diversity, climate velocity, and range position.
```{r modres}

# Summary of R2:
resR2_ok

# Model coefficients:
rownames(res_ok)=NULL
res_ok
```

## Reweighting
If we re scale the weights to N rather than 1, the fit is no longer singular. Still, these weights are inappropriate.
```{r reweight}
mydatatogo2$weight_obs_spp_2<-mydatatogo2$weight_obs_spp*nrow(mydatatogo2)

reweighted <- glmmTMB(gam_velxGDxedge1,
                family = Gamma(link = 'log'),
                weights = mydatatogo2$weight_obs_spp_2,
                REML = FALSE,
                data = mydatatogo2)

summary(reweighted) 
check_singularity(reweighted)
MuMIn::r.squaredGLMM(reweighted)
```

## Weighting simulation
The effects of summing the weights to 1 can also be seen in a simple simulation
```{r reweight}
group <- rep(1:5, each=20)
x     <- rnorm(100)
y     <- rgamma(100, 2, 2/exp(1 + 0.5*x + rep(rnorm(5,0,0.6), each=20)))
dat   <- data.frame(y=y, x=x, g=factor(group))


# weights sum to 1... each weight approx 1/N, collapses likelihood 
m1 <- glmmTMB(y~x+(1|g), data=dat, family=Gamma(link="log"), weights=rep(1/100,100), REML=FALSE)

# weights sum to N... absolute scale, RE variance correctly estimated
m2 <- glmmTMB(y~x+(1|g), data=dat, family=Gamma(link="log"), weights=rep(1,100), REML=FALSE)

# no weights matches m2
m3 <- glmmTMB(y~x+(1|g), data=dat, family=Gamma(link="log"), REML=FALSE)


summary(m1)  # broken
summary(m2)  # fine
summary(m3)  # fine
```


## Adjusted model
We fit the same model nesting a species random effect within class and removing weights:
```{r altmodel}

gam_velxGDxedge1.alt <- as.formula(
    "SHIFT_abs ~ scale(vel_abs) * scale(GD) * Param + 
    LogNtempUnits + LogExtent + ContinuousGrain + PrAb + Quality + 
    (1|Class/spp)")

alt.gam <- glmmTMB(gam_velxGDxedge1.alt, 
                     family = Gamma(link = "log"),
                     #weights = weight_obs_spp,
                     REML=F,
                     data = mydatatogo2)

```

The model is not singular with both random effects contributing. With lmerTest, there are no significant effects, but we explore more robust methods below.
```{r altsummary}
summary(alt.gam)
check_singularity(alt.gam)
MuMIn::r.squaredGLMM(alt.gam)
```

## Parametric LRT bootstrap
We use a parametric LRT bootstrap to assess the statistical support for the effects of genetic diversity on our alternative model. We used the simulateLRT function from the DHARMa package and tested the full model (with species random effect) against a null model without genetic diversity or any of its interactions. We ran the LRT bootstrap in a separate script (lrt-bootstrap.R) and read in the results here.

### There is no evidence that including any effect of genetic diversity improves model fit
```{r lrtbootstrapGDresults}
lrt_res<-readRDS(here("data", "lrt_bootstrap.rds"))
lrt_res
```

## Corrected bootstrap
We also evaluated support for effects of genetic diversity using our model and their bootstrap method, but corrected to appropriately sample species and then observations within species
```{r fixedbootstrap, eval=FALSE}
# cluster setup
nCPU <- 10
s1 <- 69
nB <- 1000
my.cluster5 <- parallel::makeCluster(nCPU, type = "PSOCK")
clusterExport(cl = my.cluster5, 
              varlist = c("s1", "nB", "glmmTMB", "mydatatogo", 
                         "gam_velxGDxedge1.alt", "check_singularity", "ranef"))
clusterEvalQ(cl = my.cluster5, {
  library(glmmTMB)
  library(MuMIn)
})
doParallel::registerDoParallel(cl = my.cluster5)

# fixed bootstrap our model
ex=pbapply::pblapply(1:nB, function(i) {
    
    gc(reset=T)
    #print(i)
    set.seed(s1+i)
    
    ## hierarchical bootstrap
    spp_boot <- sample(unique(mydatatogo$spp),
                   size = length(unique(mydatatogo$spp)),
                   replace = TRUE)

    boot_list <- vector("list", length(spp_boot))
    for(j in seq_along(spp_boot)){
        tmp <- subset(mydatatogo,
                  spp == spp_boot[j])
       tmp <- tmp[sample(1:nrow(tmp),
                      size = nrow(tmp),
                      replace = TRUE), ]
        tmp$spp <- paste0(tmp$spp[1], "_boot", j)

      boot_list[[j]] <- tmp
    }

    mydatatogo2 <- do.call(rbind, boot_list)
    
    # mydatatogoCE=subset(mydatatogo2,Param=="O")
    # x1=data.frame(table(mydatatogoCE$spp))
    # x1=subset(x1,Freq>0)
    # x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    # mydatatogoCE=merge(mydatatogoCE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    # 
    # mydatatogoLE=subset(mydatatogo2,Param=="LE")
    # x1=data.frame(table(mydatatogoLE$spp))
    # x1=subset(x1,Freq>0)
    # x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    # mydatatogoLE=merge(mydatatogoLE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    # 
    # mydatatogoTE=subset(mydatatogo2,Param=="TE")
    # x1=data.frame(table(mydatatogoTE$spp))
    # x1=subset(x1,Freq>0)
    # x1$weight_obs_spp=(1/x1$Freq)*(1/nrow(x1))
    # mydatatogoTE=merge(mydatatogoTE,x1[,c(1,3)],by.x="spp",by.y="Var1")
    # 
    # mydatatogoTE$weight_obs_spp=mydatatogoTE$weight_obs_spp*(1/3)
    # mydatatogoCE$weight_obs_spp=mydatatogoCE$weight_obs_spp*(1/3)
    # mydatatogoLE$weight_obs_spp=mydatatogoLE$weight_obs_spp*(1/3)
    # 
    # mydatatogo2=rbind(mydatatogoTE,mydatatogoCE,mydatatogoLE)
    
    gamX1 <- glmmTMB(gam_velxGDxedge1.alt, 
                     family = Gamma(link = "log"),
                     #weights = weight_obs_spp,
                     REML=F,
                     data = mydatatogo2)
    round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=round(MuMIn::r.squaredGLMM(gamX1),2)
    r2=data.frame(r2m=r2[2,1],
                  r2c=r2[2,2],
                  moy_GD=mean(mydatatogo2$GD), 
                  sd_GD=sd(mydatatogo2$GD),
                  moy_VC=mean(mydatatogo2$vel_abs), 
                  sd_VC=sd(mydatatogo2$vel_abs),
                  singu=check_singularity(gamX1)[[1]],
                  nB=i)
    
    coeff=data.frame(summary(gamX1)$coeff$cond,
                     var=row.names(summary(gamX1)$coeff$cond),
                     nB=i)
    
    randEff=as.data.frame(ranef(gamX1,condVar=T))
    randEff$nB=i
    
    return(list(r2,coeff,randEff))
    
}, cl = my.cluster5)
parallel::stopCluster(cl = my.cluster5)

saveRDS(ex, here("data", "bootstrap_fixed_raw.rds"))

mod<-"altBoot"
a=1
for(i in 1:length(mod)){
    print(a)
    r2 <- do.call(rbind, lapply(ex, `[[`, 1))
    c1 <- do.call(rbind, lapply(ex, `[[`, 2))
    r2a <- subset(r2, !is.na(r2m))
  
    #r2=read.csv2(paste0("R2_",mod[i],".csv"),sep=";",dec=".")
    #r2$nB=1:nrow(r2)
    #r2a=subset(r2,is.na(r2m)==F)
    print(paste(mod[i],': ',nrow(r2a),sep=""))
    if(nrow(r2a)>=nB){ #random selection of nB bootstrap model (criteria: model convergence)
        set.seed(s1)
        r2a=r2a[sample(1:nrow(r2a),size=nB,replace=F),]
    }
    if(nrow(r2a)<nB){
        print(paste0("sampling size is less than ",nB," bootstraps"))
    }else{
        resR2=data.frame(type="r2m",moy=mean(r2a$r2m),sd=sd(r2a$r2m),median=median(r2a$r2m),q025=quantile(r2a$r2m,probs=0.025),q975=quantile(r2a$r2m,probs=0.975))
        resR2=rbind(resR2,data.frame(type="r2c",moy=mean(r2a$r2c),sd=sd(r2a$r2c),median=median(r2a$r2c),q025=quantile(r2a$r2c,probs=0.025),q975=quantile(r2a$r2c,probs=0.975)))
        #resR2$n_sing=nrow(subset(r2a,sing==T))
        resR2$model=mod[i]
        
        #c1=read.csv2(paste("coeff_",mod[i],".csv",sep=""),sep=";",dec=".",h=T)
        #c1$nB=rep(1:nrow(r2),each=nrow(c1)/6000)
        c1=merge(c1,data.frame(nB=r2a$nB),by.x="nB",by.y="nB")
        rr3=c1
        
        res=data.frame(var=names(tapply(rr3$Estimate,rr3$var,mean)),moy=tapply(rr3$Estimate,rr3$var,mean),sd=tapply(rr3$Estimate,rr3$var,sd),median=tapply(rr3$Estimate,rr3$var,median),q025=tapply(rr3$Estimate,rr3$var,quantile,probs=0.025),
                       p975=tapply(rr3$Estimate,rr3$var,quantile,probs=0.975), pv.inf0=tapply(rr3$Estimate,rr3$var,test2,mu=0),pv.sup0=tapply(rr3$Estimate,rr3$var,test1,mu=0))
        res$model=mod[i]
        if(a==1){
            resR2_ok=resR2
            res_ok=res
        }
        else{
            resR2_ok=rbind(resR2_ok,resR2)
            res_ok=rbind(res_ok,res)
        }
        
        a=a+1
    }
}

saveRDS(list(resR2_ok = resR2_ok, res_ok = res_ok), here("data", "bootstrap_fixed.rds"))
```

### No support of any GD effects using our model and a proper bootstrap
```{r fixed bootstrapresults}
bootstrap_fixed <- readRDS(here("data", "bootstrap_fixed.rds"))
resR2_ok <- bootstrap_fixed$resR2_ok
res_ok <- bootstrap_fixed$res_ok

# summary of r2
resR2_ok

# summary of coefficients
res_ok
```